---
title: "John Askham"
format: html
editor: visual
---

## 

```{r}
#| label: setup
#| echo: fenced
#| message: false

# load packages here
library(readxl)
library(knitr)
library(tidyverse)
library(moderndive)
library(waldo)
# set theme for plotting
# theme_set(theme_minimal()) 

```

> 4a) Read the provided Excel file "Folkatal_1985_2025.xlsx" into the object `folkatal`, skipping the metadata line. Display the object as simply as possible (do not use `kable()`).

```{r}
#| label: 4a
#| echo: fenced

# code here

folkatal <- read_excel("data/Folkatal_1985_2025.xlsx", skip = 1)

folkatal
```

> 4b) Note that many cells in column `Kyn` are `NA`. This is because R assumes empty cells in the Excel file to be `NA` (missing values). In this case, however, these cells are empty because the data provider only fills the first cell of the column until the value changes. Therefore, these are not truly missing values and should be filled. Pipe the imported data `folkatal` into a fill function from the `tidyr` package to fill the `NA` cells of column `Kyn` with the correct values. Save your results in an updated object `folkatal` and display it.

```{r}
#| label: 4b
#| echo: fenced

# code here

folkatal <- folkatal |>
  fill(Kyn, .direction = "down")

kable(folkatal)

```

> 4c) Note that the column `Aldur` in `folkatal` is of type `chr` and contains numbers and text. Use a pipe and `mutate()` together with a string replacement function from `stringr` to modify the column so that it contains only the numbers. Save your results in an updated object `folkatal` and display it. Check that the type of column `Aldur` is still `chr` and do not change it here.

```{r}
#| label: 4c
#| echo: fenced



# code here
folkatal <- folkatal |>
  mutate(Aldur = str_remove(Aldur, " ĂĄr"))

tibble(folkatal)

```

> > 4d) Next, consider the many columns that contain years as column names, of both type `chr` and `dbl`. Tidy the `folkatal` object by using a pivot function from the `tidyr` package, ensuring that you get two new columns, `Ăr` and `Antal`, both of type `integer`. Note the warnings: "Warning: NAs introduced by coercion", that you will handle in the next question. Save your results in a new object `tidy_folkatal` and display it.

```{r}
#| label: 4d
#| echo: fenced

# code here

# Eg eri ikki 100% sikkur Ă­ at eg havi gjĂ¸rt hettar rĂŚtt
# Men eg fĂĄi taĂ° warnng Ă­ uppgĂĄvu orĂ°ingin sigur skal koma.
tidy_folkatal <- folkatal |>
  pivot_longer(
    cols = -c(Kyn, Aldur),
    names_to = "Ăr",
    values_to = "Antal",
    values_transform = list(Antal = as.character)) |>
    mutate(
    Ăr = as.integer(Ăr),
    Antal = as.integer(Antal)
  )

tidy_folkatal

```

> 4e) Note that some cells initially contained a dash `-` used by the data provider to denote "Veruliga null (ikki rundaĂ°)", i.e., zero values. In 4d), the cells containing a dash were automatically coerced to `NA`, which produced the warnings: "Warning: NAs introduced by coercion". Here, however, these values should be set to zero. Therefore, use a replacement function from `tidyr` to replace all `NA` in column `Antal` with `0`. Save the result in an updated `tidy_folkatal` and display it.

```{r}
#| label: 4e
#| echo: fenced

# code here
# Did this to double check the amount of "NA" within the table
sum(is.na(tidy_folkatal$Antal))

tidy_folkatal |> 
  filter(is.na(Antal))


tidy_folkatal <- tidy_folkatal |>
  mutate(Antal = replace_na(Antal, 0))


sum(is.na(tidy_folkatal$Antal))

tidy_folkatal |> 
  filter(is.na(Antal))

tidy_folkatal
```

> 4f1) For pedagogical reasons, we imported and transformed the data in several steps given in questions 4a)âe). However, we could have done it all in a single pipe sentence. Create one pipe sentence that combines these steps. Save the output in `folk` and display it.

```{r}
#| label: 4f1
#| echo: fenced

# code here



folk <- read_excel("data/Folkatal_1985_2025.xlsx", skip = 1) |>
  fill(Kyn, .direction = "down") |>
  mutate(Aldur = str_remove(Aldur, " ĂĄr")) |>
  pivot_longer(
    cols = -c(Kyn, Aldur),
    names_to = "Ăr",
    values_to = "Antal",
    values_transform = list(Antal = as.character)
  ) |>
  mutate(
    Ăr = as.integer(Ăr),
    Antal = as.integer(Antal),
    Antal = replace_na(Antal, 0)
  )

folk

```

> 4f2) Ensure that your `tidy_folkatal` and `folk` objects are exactly equal. This can be checked with `identical(x, y)` from base R, which returns`TRUE` if `x` and `y` are exactly equal or `FALSE` if not. If you get `FALSE`, use, for example `all.equal(x, y)` from base R, which returns `TRUE` if `x` and `y` are nearly equal and reports differences if not. Then, correct your code in 4f1) until `identical()` in 4f2) returns `TRUE`. A more modern alternative is to use `waldo::compare()` from the `waldo` package to compare and to report differences, if any. Then, fix your code in 4f1) until `identical()` returns `TRUE` and `waldo::compare()` reports "No differences".

```{r}
#| label: 4f2
#| echo: fenced

# code here


identical(tidy_folkatal, folk)

all.equal(tidy_folkatal, folk)

waldo::compare(tidy_folkatal, folk)
```

## Exercise 5

> In this exercise, the variables are in the Faroese language, and you can also use Faroese for the plot titles.

> 5a) Summarise `folk` grouped by `Ăr` to find the total population `FĂłlkatal` for each year. Save the result in `arligt_folkatal` and display it.

```{r}
#| label: 5a
#| echo: fenced

# code here
```

> 5b) Use `ggplot()` to create a linegraph of `FĂłlkatal` versus `Ăr` in `arligt_folkatal`. Also, include the data points.

```{r}
#| label: 5b
#| echo: fenced

# code here
```

> 5c) Summarise `folk` grouped by `Ăr` and `Kyn` to find the total population `FĂłlkatal` for each `Ăr` and for each `Kyn`. Save the result in `kyn_folkatal` and display it.

```{r}
#| label: 5c
#| echo: fenced

# code here
```

> 5d) Use `ggplot()` to create a linegraph of `FĂłlkatal` versus `Ăr` in `kyn_folkatal` so that you get two curves - one for each `Kyn`. Also, include the data points for each `Kyn`. To obtain a single legend, map the color of the lines to `Kyn`, and map both the color and shape of the points to `Kyn`. Use `scale_color_manual()` to set the color for "Menn" to "blue" and for "Kvinnur" to "red".

```{r}
#| label: 5d
#| echo: fenced

# code here

```

> 5e) Reuse the plot code you created in 5d). Add a similar curve of `FĂłlkatal` versus `Ăr` from `arligt_folkatal` so that you get three curves in total - the previous two curves of `FĂłlkatal` per `Kyn`, described in the legend, and one additional curve of `FĂłlkatal` for the total population. We also want to include the third curve in the same legend. However, there is no column `Kyn` in `arligt_folkatal`. Therefore, for the curve of `FĂłlkatal` representing the total population, you cannot map color or shape to `Kyn`. Instead, map color and shape to the text "BĂŚĂ°i" to create an additional entry in the legend describing the total population. Modify `scale_color_manual()` in the plot code to define the colors for "BĂŚĂ°i", "Menn", and "Kvinnur" as "black", "blue", and "red", respectively.

```{r}
#| label: 5e
#| echo: fenced

# code here
```

> 5f1) `kyn_folkatal` has the columns `Ăr, Kyn, FĂłlkatal`, while `arligt_folkatal` only has the columns `Ăr, FĂłlkatal`. Pipe `arligt_folkatal` into the same structure as `kyn_folkatal`. Add a column `Kyn` with the value "BĂŚĂ°i", reorder the columns, and bind it to `kyn_folkatal`. Save the output as `alt_folkatal` and display it. Check that it has the columns `Ăr, Kyn, FĂłlkatal` with data types `int, chr, int`, respectively, and that the number of rows equals the number of rows in `kyn_folkatal` plus the number of rows in `arligt_folkatal`.

```{r}
#| label: 5f1
#| echo: fenced

# code here
```

> 5f2) Simplify your plot code from 5e) by using only `alt_folkatal` as the data source to create a plot identical to the one in 5e).

```{r}
#| label: 5f2
#| echo: fenced

# code here

```

> 5f3) Now, we want to change the order of the legend text. Convert the `Kyn` column in `alt_folkatal` to a factor variable with levels "BĂŚĂ°i", "Menn", "Kvinnur". Save the changes in an updated `alt_folkatal` and display it. Check that it now has the columns `Ăr, Kyn, FĂłlkatal` with data types `int, fctr, int`, respectively.

```{r}
#| label: 5f3
#| echo: fenced

# code here
```

> 5f4) Run the exact same code as in 5f2) using the updated `alt_folkatal`. Note what changed as compared to the plot in 5f2)?

```{r}
#| label: 5f4
#| echo: fenced

# code here

```
