---
title: "John Askham"
format: html
editor: visual
---

## 

```{r}
#| label: setup
#| echo: fenced
#| message: false

# load packages here
library(readxl)
library(knitr)
library(tidyverse)
library(moderndive)
library(waldo)
# set theme for plotting
# theme_set(theme_minimal()) 

```

> 4a) Read the provided Excel file "Folkatal_1985_2025.xlsx" into the object `folkatal`, skipping the metadata line. Display the object as simply as possible (do not use `kable()`).

```{r}
#| label: 4a
#| echo: fenced

# code here

folkatal <- read_excel("data/Folkatal_1985_2025.xlsx", skip = 1)

folkatal
```

> 4b) Note that many cells in column `Kyn` are `NA`. This is because R assumes empty cells in the Excel file to be `NA` (missing values). In this case, however, these cells are empty because the data provider only fills the first cell of the column until the value changes. Therefore, these are not truly missing values and should be filled. Pipe the imported data `folkatal` into a fill function from the `tidyr` package to fill the `NA` cells of column `Kyn` with the correct values. Save your results in an updated object `folkatal` and display it.

```{r}
#| label: 4b
#| echo: fenced

# code here

folkatal <- folkatal |>
  fill(Kyn, .direction = "down")

kable(folkatal)

```

> 4c) Note that the column `Aldur` in `folkatal` is of type `chr` and contains numbers and text. Use a pipe and `mutate()` together with a string replacement function from `stringr` to modify the column so that it contains only the numbers. Save your results in an updated object `folkatal` and display it. Check that the type of column `Aldur` is still `chr` and do not change it here.

```{r}
#| label: 4c
#| echo: fenced



# code here
folkatal <- folkatal |>
  mutate(Aldur = str_remove(Aldur, " ár"))

tibble(folkatal)

```

> > 4d) Next, consider the many columns that contain years as column names, of both type `chr` and `dbl`. Tidy the `folkatal` object by using a pivot function from the `tidyr` package, ensuring that you get two new columns, `Ár` and `Antal`, both of type `integer`. Note the warnings: "Warning: NAs introduced by coercion", that you will handle in the next question. Save your results in a new object `tidy_folkatal` and display it.

```{r}
#| label: 4d
#| echo: fenced

# code here

# Eg eri ikki 100% sikkur í at eg havi gjørt hettar rætt
# Men eg fái tað warnng í uppgávu orðingin sigur skal koma.
tidy_folkatal <- folkatal |>
  pivot_longer(
    cols = -c(Kyn, Aldur),
    names_to = "Ár",
    values_to = "Antal",
    values_transform = list(Antal = as.character)) |>
    mutate(
    År = as.integer(Ár),
    Antal = as.integer(Antal)
  )

tidy_folkatal

```

> 4e) Note that some cells initially contained a dash `-` used by the data provider to denote "Veruliga null (ikki rundað)", i.e., zero values. In 4d), the cells containing a dash were automatically coerced to `NA`, which produced the warnings: "Warning: NAs introduced by coercion". Here, however, these values should be set to zero. Therefore, use a replacement function from `tidyr` to replace all `NA` in column `Antal` with `0`. Save the result in an updated `tidy_folkatal` and display it.

```{r}
#| label: 4e
#| echo: fenced

# code here
# Did this to double check the amount of "NA" within the table
sum(is.na(tidy_folkatal$Antal))

tidy_folkatal |> 
  filter(is.na(Antal))


tidy_folkatal <- tidy_folkatal |>
  mutate(Antal = replace_na(Antal, 0))


sum(is.na(tidy_folkatal$Antal))

tidy_folkatal |> 
  filter(is.na(Antal))

tidy_folkatal
```

> 4f1) For pedagogical reasons, we imported and transformed the data in several steps given in questions 4a)–e). However, we could have done it all in a single pipe sentence. Create one pipe sentence that combines these steps. Save the output in `folk` and display it.

```{r}
#| label: 4f1
#| echo: fenced

# code here



folk <- read_excel("data/Folkatal_1985_2025.xlsx", skip = 1) |>
  fill(Kyn, .direction = "down") |>
  mutate(Aldur = str_remove(Aldur, " ár")) |>
  pivot_longer(
    cols = -c(Kyn, Aldur),
    names_to = "Ár",
    values_to = "Antal",
    values_transform = list(Antal = as.character)
  ) |>
  mutate(
    År = as.integer(Ár),
    Antal = as.integer(Antal),
    Antal = replace_na(Antal, 0)
  )

folk

```

> 4f2) Ensure that your `tidy_folkatal` and `folk` objects are exactly equal. This can be checked with `identical(x, y)` from base R, which returns`TRUE` if `x` and `y` are exactly equal or `FALSE` if not. If you get `FALSE`, use, for example `all.equal(x, y)` from base R, which returns `TRUE` if `x` and `y` are nearly equal and reports differences if not. Then, correct your code in 4f1) until `identical()` in 4f2) returns `TRUE`. A more modern alternative is to use `waldo::compare()` from the `waldo` package to compare and to report differences, if any. Then, fix your code in 4f1) until `identical()` returns `TRUE` and `waldo::compare()` reports "No differences".

```{r}
#| label: 4f2
#| echo: fenced

# code here


identical(tidy_folkatal, folk)

all.equal(tidy_folkatal, folk)

waldo::compare(tidy_folkatal, folk)
```

```{r}

```

```{r}

```

```{r}

```
